Scheduling and Amortization for MPC

### 2021-12-31 paper_SIMD.tex
minor edit about Vassilis' suggestion about cost
minor edit
Minor edits to Extension of Basic Vectorization

\begin{enumerate}
\item All statements in the program execute using the same protocol (sharing). That is, there is no share conversion. \ishaq{We should say somewhere (in a footnote?) that we know this is not ideal, otherwise reviewers will complain.}
\item All MPC instructions have the same unit cost, 1 unit. \ishaq{Don't we need to distinguish between local (cheap) vs. interactive (expensive) instructions? I don't see how we could assume this.}
\item There is unlimited bandwidth---i.e., a single MPC-instruction costs as much as $N$ amortized instructions, namely 1 unit.\ishaq{This is fair for theoretical model, although in some real life applications this is not true, e.g. L2 norm of large vectors.}
\item MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption, 
2 MUL instructions scheduled in parallel benefit from amortization and cost 1, however a MUL and a MUX instructions scheduled 
in parallel still cost 2. \ishaq{This is not true, can we still assume it?} 
\end{enumerate}

@@

Array writes limit vectorization as they introduce sometimes infeasible loop-carried dependencies. Consider the following example: \ana{Have to add citation to Aiken's paper}

@@

\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{A}_0$ = $\phi(\texttt{A},\texttt{A}_1)$ 
\STATE 2. $\texttt{B}_0$ = $\phi(\texttt{B},\texttt{B}_1)$ 
\STATE 3. $\texttt{C}_0$ = $\phi(\texttt{C},\texttt{C}_1)$ 
\STATE 4. $\texttt{D}_0$ = $\phi(\texttt{D},\texttt{D}_1)$ 
\STATE 5. $\texttt{A}_1$ = $\texttt{A}_0$ 
\STATE 6. $\texttt{A}_1$[i] = $\texttt{B}_0$[i] + 10;
\STATE 7. $\texttt{B}_1$ = $\texttt{B}_0$ 
\STATE 8. $\texttt{B}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 9. $\texttt{C}_1$ = $\texttt{C}_0$ 
\STATE 10. $\texttt{C}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 11. $\texttt{D}_1$ = $\texttt{D}_0$ 
\STATE 12. $\texttt{D}_1$[i] = $\texttt{B}_1$[i] * $\texttt{C}_1$[i];
\end{algorithmic}

@@

\ana{TODO: We may remove phi-nodes that end up without incoming back-edges. This can free some cycles that we'd waste on bookkeeping arrays.}

@@

%\ana{Clarify why this helps. Still work in progress.}
This is to reduce the dimensionality of simd-ified computation. Technically, $\texttt{A}_j = \emph{MUX}(..., \texttt{A}_k, \texttt{A}_l)$ is 
a simdified operation that can be carried out in parallel "in one round". However, particularly when \texttt{A} is a multi-dimensional array, there is
substantial increase in the size of the arrays (vectors) we send to SIMD operations. Refining to an update to a specific index would reduce 
the size of those vectors. Note that this is a heuristic that handles a common case, but not all cases of array updates. 
\ana{It may actually be handling \_all\_ cases, with the restriction to canonical updates.}

\begin{algorithmic}
\FOR {each $\mathit{stmt}$: $\texttt{A}_j = \emph{MUX}(c, \texttt{A}_k, \texttt{A}_l)$ in the MPC-source seq.}
\STATE $i_1 = \mathit{find\_update}(A_k)$ \COMMENT { Is null when $A_k = \phi(...)$ }
\STATE $i_2 = \mathit{find\_update}(A_l)$ \COMMENT { Is null when $A_l = \phi(...)$ }
\IF {$i_1 == i_2$ or $i_1$ is null or $i_2$ is null} \STATE \COMMENT{ With our restrictions on writes we must have $i_1 = i_2$. } 
\STATE replace $\mathit{stmt}$ with \\
\STATE $\texttt{A}_j = \texttt{A}_{j-1}$; $\texttt{A}_j[i_1] = \emph{MUX}(c, \texttt{A}_k[i_1], \texttt{A}_l[i_1])$
%\ELSIF {$i_1 \ncong i_2$}  
%\STATE replace $\mathit{stmt}$ with \\
%\STATE $\texttt{A}_j = \texttt{A}_k$; $\texttt{A}_j[i_1] = \emph{MUX}(c, \texttt{A}_k[i_1], \texttt{A}_l[i_1])$
\ELSE \STATE
\STATE $\mathit{stmt}$ stays as is
\ENDIF
\ENDFOR
\end{algorithmic}

@@

\item inner-to-outer $X \rightarrow Y$. Let $A$ be the array defined at $X$. If the dimensionality $A$ is greater than its canonical dimensionality, then add \emph{drop\_dim(...)} at $Y$, as in Basic Vectorization. Otherwise, do nothing.

@@

After removal of infeasible edges, the Aiken's array write example will be (roughly) as follows:

\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{D}_0$ = $\phi(\texttt{D},\texttt{D}_1)$ 
\STATE 2. $\texttt{A}$[i] = $\texttt{B}$[i] + 10
\STATE 3. $\texttt{B}$[i] = $\texttt{A}$[i] * $\texttt{D}_0$[i-1]
\STATE 4. $\texttt{C}$[i] = $\texttt{A}$[i] * $\texttt{D}_0$[i-1]
\STATE 5. $\texttt{D}_1$ = $\texttt{D}_0$ 
\STATE 6. $\texttt{D}_1$[i] = $\texttt{B}$[i] * $\texttt{C}$[i]
\end{algorithmic}

@@

After Phase 1 of Basic vectorization we have the following code:
\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{D}_0$ = $\phi(\texttt{D},\texttt{D}_1)$ 
\STATE 2. $\texttt{A}[\stackrel{\rightarrow}{i}] = \texttt{B}[\stackrel{\rightarrow}i] + [10,...]$
\STATE 3. $\texttt{B}[\stackrel{\rightarrow}i] = \texttt{A}[\stackrel{\rightarrow}i] * \texttt{D}_0[\stackrel{\rightarrow}{i-1}]$
\STATE 4. $\texttt{C}[\stackrel{\rightarrow}i] = \texttt{A}[\stackrel{\rightarrow}i] * \texttt{D}_0[\stackrel{\rightarrow}{i-1}]$
\STATE 5. $\texttt{D}_1$ = $\texttt{D}_0$ 
\STATE 6. $\texttt{D}_1[\stackrel{\rightarrow}i] = \texttt{B}[\stackrel{\rightarrow}i] * \texttt{C}[\stackrel{\rightarrow}i]$
\end{algorithmic}

In Phase 2 we get rid of the arrows on statements 3, 4, and 6 as they are not vectorizable and in Phase 3 we add SIMD appropriately:

\begin{algorithmic}
\STATE 1. $\texttt{A}[\stackrel{\rightarrow}{i}] = \mathit{ADD\_SIMD}(\texttt{B}[\stackrel{\rightarrow}i],[10,...])$ \COMMENT{Fully vectorized, size N.} 
\STATE FOR i=0; i<N; i++; \COMMENT{ MOTION loop }
\STATE 2. $\texttt{D}_0$ = $\phi(\texttt{D},\texttt{D}_1)$ 
\STATE 3. $\texttt{B}[i] = \mathit{MUL}(\texttt{A}[i], \texttt{D}_0[{i-1}])$
\STATE 4. $\texttt{C}[i] = \mathit{MUL}(\texttt{A}[i], \texttt{D}_0[{i-1}])$
\STATE 5. $\texttt{D}_1$ = $\texttt{D}_0$
\STATE 6. $\texttt{D}_1[i] = \mathit{MUL}(\texttt{B}[i], \texttt{C}[i])$
\end{algorithmic}

@@

Consider MPC-source of Histogram after removal of infeasible edges and MUX refinement:

@@

\STATE res!1 = []
\STATE for i: plaintext in range(0, num\_bins):
\STATE \hspace{0.25cm} res!2 = $\phi$(res!1, res!3)
\STATE  \hspace{0.25cm} ... 
\STATE res!4 = res!2 \COMMENT{Added due to the "mixed" def-use edge.} 
\STATE for i: plaintext in range(0, num\_bins):
%\STATE \hspace{0.25cm}        result!4 = $\phi$(result!2, result!5)
\STATE \hspace{0.25cm}        for j: plaintext in range(0, N):
\STATE \hspace{0.5cm}            res!5 = $\phi$(res!4, res!7)
\STATE \hspace{0.5cm}            !2!3 = (A[j] == i)
\STATE \hspace{0.5cm}            !3!3 = (res!5[i] + B[j])
\STATE \hspace{0.5cm}            res!6 = res!5
\STATE \hspace{0.5cm}            res!6[i] = !3!3 
\STATE \hspace{0.5cm}            res!7 = res!6 
\STATE \hspace{0.5cm}            res!7[i] = MUX(!2!3, res!6[i], res!5[i])

>>>>

\begin{enumerate}
\item All statements in the program execute using the same protocol (sharing). That is, there is no share conversion. \ishaq{We should say somewhere (in a footnote?) that we know this is not ideal, otherwise reviewers will complain.}
\item All MPC instructions have the same unit cost, 1 unit. \ishaq{Don't we need to distinguish between local (cheap) vs. interactive (expensive) instructions? I don't see how we could assume this. -- Past suggestion from Vassilis: Ideally we should assume cost to be a convex function.}
\item There is unlimited bandwidth---i.e., a single MPC-instruction costs as much as $N$ amortized instructions, namely 1 unit.\ishaq{This is fair for theoretical model, although in some real life applications this is not true, e.g. L2 norm of large vectors.}
\item MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption, 
2 MUL instructions scheduled in parallel benefit from amortization and cost 1, however a MUL and a MUX instructions scheduled 
in parallel still cost 2. \ishaq{This is not true, can we still assume it? -- YES.} 
\end{enumerate}

@@

Array writes limit vectorization as they sometimes introduce infeasible loop-carried dependencies. Consider the following example: \ana{Have to add citation to Aiken's paper}

@@

\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{A}_0$ = $\phi(\texttt{A},\texttt{A}_1)$ 
\STATE 2. $\texttt{B}_0$ = $\phi(\texttt{B},\texttt{B}_1)$ 
\STATE 3. $\texttt{C}_0$ = $\phi(\texttt{C},\texttt{C}_1)$ 
\STATE 4. $\texttt{D}_0$ = $\phi(\texttt{D},\texttt{D}_1)$ 
\STATE 5. $\texttt{A}_1$ = $\texttt{A}_0$; $\texttt{A}_1$[i] = $\texttt{B}_0$[i] + 10; \COMMENT{equiv. to Update}
\STATE 6. $\texttt{B}_1$ = $\texttt{B}_0$; $\texttt{B}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 7. $\texttt{C}_1$ = $\texttt{C}_0$; $\texttt{C}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 8. $\texttt{D}_1$ = $\texttt{D}_0$; $\texttt{D}_1$[i] = $\texttt{B}_1$[i] * $\texttt{C}_1$[i];
\end{algorithmic}

@@

\ana{TODO: We need to remove phi-nodes that end up without incoming back-edges.}

The final step is to remove array phi-nodes that are left without incoming back edges after removal of infeasible edges.
Clearly, such a phi-node is redundant when there are no loop-carried dependences that are due to writes to the corresponding array.
In our running example, the phi-nodes for \texttt{A}, \texttt{B}, and \texttt{C}, are all redundant.

Removing redundant phi-nodes is as follows. 
Denote a redundant phi-node by $\texttt{A}_1 = \phi(\texttt{A}_0,\texttt{A}_k)$. $\texttt{A}_0$ is the array at the beginning of the loop
and $\texttt{A}_k$ is the array at the end of an iteration (including the last iteration). For each outer-to-inner or same-level def-use edge 
that originates at the node, replace $\texttt{A}_1$ at the use node by $\texttt{A}_0$. This is correct, i.e., it is guaranteed that the use 
references $\texttt{A}_0$; had the use been referencing an array modified in an earlier iteration, the back-edge to the phi-node would
not have been removed and the phi-node would not have been classified as redundant. 
For each inner-to-outer def-use edge, replace $\texttt{A}_1$ by $\texttt{A}_k$. In this case, the use references the array
that results from the computation in the loop.

Our running example is transformed as follows. Removing the phi-nodes for \texttt{A}, \texttt{B}, and \texttt{C} transforms the 
replaces the references to $\texttt{A}_0$, $\texttt{B}_0$, and $\texttt{C}_0$, with references to $\texttt{A}$, $\texttt{B}$, and $\texttt{C}$
respectively, i.e., the arrays before the loop.

\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{D}_0$ = $\phi(\texttt{D},\texttt{D}_1)$ 
\STATE 2. $\texttt{A}_1$ = $\texttt{A}$; $\texttt{A}_1$[i] = $\texttt{B}$[i] + 10;
\STATE 3. $\texttt{B}_1$ = $\texttt{B}$; $\texttt{B}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 4. $\texttt{C}_1$ = $\texttt{C}$; $\texttt{C}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 5. $\texttt{D}_1$ = $\texttt{D}_0$; $\texttt{D}_1$[i] = $\texttt{B}_1$[i] * $\texttt{C}_1$[i];
\end{algorithmic}

@@

%\ana{Clarify why this helps. Still work in progress.}
This is to reduce the dimensionality of simd-ified computation. Technically, $\texttt{A}_j = \emph{MUX}(..., \texttt{A}_k, \texttt{A}_l)$ is 
a simdified operation that can be carried out in parallel "in one round". However, particularly when \texttt{A} is a multi-dimensional array, there is
substantial increase in the size of the arrays (vectors) we send to SIMD operations. Refining to an update to a specific index would reduce 
the size of those vectors. Note that this is a heuristic that handles a common case, but not all cases of array updates. 
\ana{TODO: Simplify this algorithm, taking into account the restriction to canonical updates. It should be handling \_all\_ cases.}

\begin{algorithmic}
\FOR {each $\mathit{stmt}$: $\texttt{A}_j = \emph{MUX}(c, \texttt{A}_k, \texttt{A}_l)$ in the MPC-source seq.}
\STATE $i_1 = \mathit{find\_update}(A_k)$ \COMMENT { Is null when $A_k = \phi(...)$ }
\STATE $i_2 = \mathit{find\_update}(A_l)$ \COMMENT { Is null when $A_l = \phi(...)$ }
\IF {$i_1 == i_2$ or $i_1$ is null or $i_2$ is null} \STATE \COMMENT{ With our restrictions on writes we must have $i_1 = i_2$. } 
\STATE replace $\mathit{stmt}$ with \\
\STATE $\texttt{A}_j = \texttt{A}_{j-1}$; $\texttt{A}_j[i_1] = \emph{MUX}(c, \texttt{A}_k[i_1], \texttt{A}_l[i_1])$
%\ELSIF {$i_1 \ncong i_2$}  
%\STATE replace $\mathit{stmt}$ with \\
%\STATE $\texttt{A}_j = \texttt{A}_k$; $\texttt{A}_j[i_1] = \emph{MUX}(c, \texttt{A}_k[i_1], \texttt{A}_l[i_1])$
\ELSE 
\STATE $\mathit{stmt}$ stays as is
\ENDIF
\ENDFOR
\end{algorithmic}

@@

\item inner-to-outer $X \rightarrow Y$. Let $A$ be the array defined at $X$. If the dimensionality of $A$ is greater than its canonical dimensionality, then add \emph{drop\_dim(...)} at $Y$, as in Basic Vectorization. Otherwise, do nothing.

@@

Recall that after removal of infeasible edges and redundant phi-nodes, the Aiken's array write example will be (roughly) as follows:

\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{D}_0$ = $\phi(\texttt{D},\texttt{D}_1)$ 
\STATE 2. $\texttt{A}_1$ = $\texttt{A}$; $\texttt{A}_1$[i] = $\texttt{B}$[i] + 10;
\STATE 3. $\texttt{B}_1$ = $\texttt{B}$; $\texttt{B}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 4. $\texttt{C}_1$ = $\texttt{C}$; $\texttt{C}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 5. $\texttt{D}_1$ = $\texttt{D}_0$; $\texttt{D}_1$[i] = $\texttt{B}_1$[i] * $\texttt{C}_1$[i];
\end{algorithmic}

@@

After Phase 1 of Basic vectorization we have the following code:
\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{D}_0$ = $\phi(\texttt{D},\texttt{D}_1)$ 
\STATE 2. $\texttt{A}_1$ = $\texttt{A}$; $\texttt{A}_1[\stackrel{\rightarrow}{i}] = \texttt{B}[\stackrel{\rightarrow}i] + [10,...]$
\STATE 3. $\texttt{B}_1$ = $\texttt{B}$; $\texttt{B}_1[\stackrel{\rightarrow}i] = \texttt{A}_1[\stackrel{\rightarrow}i] * \texttt{D}_0[\stackrel{\rightarrow}{i-1}]$
\STATE 4. $\texttt{C}_1$ = $\texttt{C}$; $\texttt{C}_1[\stackrel{\rightarrow}i] = \texttt{A}_1[\stackrel{\rightarrow}i] * \texttt{D}_0[\stackrel{\rightarrow}{i-1}]$
\STATE 5. $\texttt{D}_1$ = $\texttt{D}_0$; $\texttt{D}_1[\stackrel{\rightarrow}i] = \texttt{B}_1[\stackrel{\rightarrow}i] * \texttt{C}_1[\stackrel{\rightarrow}i]$
\end{algorithmic}

In Phase 2 we get rid of the arrows on statements 3, 4, and 5 as they are not vectorizable and in Phase 3 we add SIMD appropriately:

\begin{algorithmic}
\STATE 1. $\texttt{A}_1$ = $\texttt{A}$; $\texttt{A}_1[\stackrel{\rightarrow}{i}] = \mathit{ADD\_SIMD}(\texttt{B}[\stackrel{\rightarrow}i],[10,...])$ \COMMENT{Fully vectorized, size N.} 
\STATE FOR i=0; i<N; i++; \COMMENT{ MOTION loop }
\STATE 2. $\texttt{D}_0$ = $\phi(\texttt{D},\texttt{D}_1)$ 
\STATE 3. $\texttt{B}_1$ = $\texttt{B}$; $\texttt{B}_1[i] = \mathit{MUL}(\texttt{A}_1[i], \texttt{D}_0[{i-1}])$
\STATE 4. $\texttt{C}_1$ = $\texttt{C}$; $\texttt{C}_1[i] = \mathit{MUL}(\texttt{A}_1[i], \texttt{D}_0[{i-1}])$
\STATE 5. $\texttt{D}_1$ = $\texttt{D}_0$; $\texttt{D}_1[i] = \mathit{MUL}(\texttt{B}_1[i], \texttt{C}_1[i])$
\end{algorithmic}

@@

Consider MPC-source of Histogram after removal of infeasible edges and redundant phi-nodes, and MUX refinement:

@@

\STATE res!1 = []
\STATE for i: plaintext in range(0, num\_bins):
\STATE \hspace{0.25cm} res!2 = $\phi$(res!1, res!3)
\STATE  \hspace{0.25cm} ... 
\STATE res!4 = res!2 \COMMENT{Added due to the "mixed" def-use edge.} 
\STATE for i: plaintext in range(0, num\_bins):
%\STATE \hspace{0.25cm}        result!4 = $\phi$(result!2, result!5)
\STATE \hspace{0.25cm}        for j: plaintext in range(0, N):
\STATE \hspace{0.5cm}            res!5 = $\phi$(res!4, res!7)
\STATE \hspace{0.5cm}            !2!3 = (A[j] == i)
\STATE \hspace{0.5cm}            !3!3 = (res!5[i] + B[j])
\STATE \hspace{0.5cm}            res!6 = res!5; res!6[i] = !3!3 
\STATE \hspace{0.5cm}            res!7 = res!6; res!7[i] = MUX(!2!3, res!6[i], res!5[i])
